Probability of causal contact between interstellar civilizations through MonteCarlo simulations

### 2020-01-13 Sec_methods.tex
ultimos retoques

Simulations are suitable tools to analyze systems that evolve with
time and involve randomness.
%
An advantage of simulations compared to theoretical approaches, is
that the former usually require less assumptions and simplifications,
and can be applied to systems where a theoretical model can not be
found.
%
In particular, many complex stochastic processes that can be described
by the evolution of the state of a system, can be efficiently modeled
with the discrete--event (hereafter, DE) simulation approach.
%
A system described with the DE paradigm is characterized by a set of
actors and events, where actors interact causally through a series of
events on a timeline and process these events in chronological order
\citep{ptolemaeus_system_2014, chung_simulation_2003,
ross_simulation_2012}.
%
Each event produces changes in the values of the variables that define
the process, and thus the corresponding change in the state of the
simulated system.
%
This method is well suited for the particular case of the diffusion of
intelligent signals in the galaxy, and allows to explore several
models easily.
%
We simulate the statistical properties of a set of points in space and
time that have a causal connection
at light speed, hereafter dubbed ''causal contact nodes'' (hereafter, \ceti{}).
%
% ability of emitting and receiving messages at
%
We chose this generic name in order to stress the fact that in this
analysis no broadcasting or lookout activities are required, 
and therefore not actual messages are considered, only the causal
contact.
%
For the special case of a fully efficient node that emits
and receives isotropically, a causal contact node can be considered as
a broadcasting station that has the ability to detect signals through
an active lookout program.
%
Also, it is worth mentioning that this is a general approach, and not
necessarily a \ceti{} is the host of an intelligent civilization.
%
It can be associated with a planet where life has developed,
became intelligent, reached the skills required to find the right
communication channel, sustain a search and establish a contact.
%
Alternative message processing entities could be considered, for
example interstellar beacons where life has ceased to exist but
continue with its emission or communication stations established by
probes or left intentionally by intelligent beings \citep[see, e.g.,
][]{peters_outer_2018, barlow_galactic_2013}.
%
In principle, these strategies could affect our results since it would
be easier to configure a cluster of \cetis{} that spread in time.
%
However, we do not consider this highly speculative alternatives at
this point.
%
% have been proposed
%
For the purpose of this analysis, only the communication capability is
relevant, since we study the causal contacts between the locations.
%
The system is defined by a number of actors that represent \cetis{} and
appear at different instants in time, generating events that produce
meaningful changes in the variables that describe the system, i.e., in
the arrangement of \cetis{} and their network of causal contacts.
%
For example, the appearance of a new \ceti{} in a region which is filled
with a signal emitted by another \ceti{}, will increase the number of
active \cetis{} and the number of pairs of \cetis{} in causal contact.
%
Assuming some simple hypotheses, the discrete events method can be
performed taking into account a small number of variables, which allow
to analyze the variation of the results in the model parameter space.

@@

\ffn{1} We illustrate in the Fig.~\ref{F_scheme} the schematic
representation of the region causally connected to the central
\ceti{}.
%
To this end, we use space--time diagrams, where time is represented on
the horizontal axis, and space is represented in the vertical axis.
%
In particular, we plot on the vertical axis the distance traveled by
a signal at light speed from the emitter.
%
Panel (a) represents the region $R$ (shaded polygon) in space and time
which is causally connected to the emitter (left vertex), following an
"A" type event in which the node acquires the communication
capability.
%
The sphere of first contact (SFC) is a sphere centered on the emitter,
represented in the Figure by the left angle of the region $R$.
%
This sphere grows until its radius reaches the D$_{max}$ distance, in
which it would be no longer detectable due to the decrease in the
power per unit area, which falls under the detectability threshold.
%
Similarly, the surface of last contact (SLC) is another sphere that
grows from a "D" event and carries the last signal produced by the
emitter.
%
The region which is causally connected to the emitter is then limited
by these two spheres, and has the shape of a sphere (before the D
event) or of a spherical shell (after the D event).
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b) of the Figure.
%
The receiver \ceti{} E$_1$ can listen signals from the emitter \ceti{}
E$_2$, from the first contact event (t=C$_{12}$) up to the last
contact event (t=B$_{12}$).
%
Similarly, the receiver \ceti{} E$_2$ can listen signals from the
emitter \ceti{} E$_1$, from t=C$_{21}$ up to t=B$_{21}$.

@@

%{{{     
In this scheme, the entire lifetime of a \ceti{} can then be
represented by a polygon, limited in the spatial axis by the maximum
distance of the signal $D_{max}$ and in the time axis by the time span
between the starting time (A event, concave vertex) and the ending
time (D event, convex vertex).
%
This representation allows to easily visualize the important events
that result from the existence and communications of two \cetis{}
(panel (b)).
%
The points in time where the \cetis{} acquire their communicating
capacity are dubbed ''Awakening'' events ($A_1$ and $A_2$ in the
Figure).
%
Similarly, the points in time where the \cetis{} loss their
communicating capacity are dubbed ''Doomsday'' events ($D_1$ and $D_2$
in the Figure).
%
The points in space--time where the first contact is produced for each
one of the \cetis{}, are defined as ''Contact'' events, shown as $C_1$
and $C_2$ in the Figure.
%
Finally, the two points where the contact is lost for each one the
\cetis{}, are denominated ''Blackout'' events, shown as $B_1$ and
$B_2$ in the Figure.
%
It is important to notice that there is a time delay between the
contact events of the two \cetis{} involved in this analysis, and also
a time delay between the two blackout events, so that the time range
when a bidirectional contact is possible occurs between the maximum
time of the contact points and the minimum time between the blackout
points.

A causal connection can be produced without the need of the two
involved locations being concurrent or active at the same time.
%
This is a fundamental property of the system which arises as a
consequence of the large spatial and temporal scales, where a message
is transmitted at light speed.
%
Although a \ceti{} could be active for enough time so as to transmit a
message at large distances, the limited power of the message and the
dilution that runs as the squared distance from the source, imposes a
detectability limit.
%
As a consequence of this limitation and of their finite lifetime,
considered as the period of time between the acquisition and loss of
communicating capacity, each \ceti{} will fill a spherical shell
region of the galaxy, limited by two concentric spherical surfaces.
%
The leading front, or surface of first contact (SFC) grows from the
central \ceti{} until it reaches the maximum distance $D_{max}$.
%
Thus, the volume reached causally by a \ceti{} is initially a growing
sphere.

Following the end of the civilization, there is still a region which
is filled with the emitted signals.
%
The trailing front, or surface of last contact (SLC) also grows from
the central \ceti{}, with a delay with respect to the SFC equivalent
to the lifetime of the \ceti{}, and produces a spherical shell region.
%
Any other \ceti{} within this region will be aware of the originating
\ceti{}, even if it has disappeared before the time of contact.
%
This region will grow if the surface of first contact has not yet
reached the maximum distance $D_{max}$, and will shrink otherwise
until the surface of last contact reaches $D_{max}$, producing as a
result the loss of all signal from the central \ceti{}.
%
In our approach, we consider a model galaxy where the width of the
disk is negligible with respect to the radius of the disk, so that the
\cetis{} are placed on a 2D plane disk.
%
In the 2D simulation only the intersection of the communicating
spherical shells with the plane of the galaxy is relevant, and produce
the corresponding circles or rings for the filled spheres or annular
regions of the spherical shells, respectively.

The initially growing communicating sphere, is shown over space--time
diagrams, where time is represented on the horizontal direction, and
space is represented in the vertical direction.
%
In the upper panel (a), the sphere is growing as the surface of first
contact has not reached the maximum distance.
%
In the middle panel (b) it has reached the maximum distance, so that
it remains at the same size.
%
After a Doomsday event (bottom panel, b), the signals can still be
observed, but the surface of last contact grows.
%
We also show in this Figure a schematic representation of two
emitters, E$_1$ and E$_2$, that reach each other at different times.
%
The time span for E$_i$ is (A$_i$ , D$_i$) , for $i = {1, 2}$.
%
Emitter $i$ can listen to emitter $j$ between C$_{ij}$ and B$_{ij}$.
%
The type and length of causal contact in both directions depend on the
distance and time lag between the awakening events, the maximum
distance that a signal can reach and the time period in which each
emitter is active.
%
\ffn{2}
%
In the Fig.~\ref{F_messages} we show a schematic representation of the
possible cases in which a \ceti{} E can be in causal contact with a
\ceti{} R.
%
The duration of the causal contact in one direction depends on several
factors, mainly the lifetimes, the time period between the A events
and the distance between the two \cetis{} involved in the causal
contact.

The temporal structure of the process is defined by two distribution
parameters that represent the mean time interval that an intelligent
civilization can emit and receive signals, and the mean time interval
between the emergence of an intelligent communicating civilization and
the next one.
%
The spatial structure of the simulation is given by the size and shape
of the Galactic Habitable Zone and the maximum distance a signal can
travel to be detected.
%
The parameters for the temporal distributions also determine the
spatial properties since the density of active \cetis{} in the galaxy
depend on these two parameters.
%
For example, a small $\tau_a$ and a large $\tau_s$ will produce a
galaxy densely populated with \cetis{}, and vice versa.
%
Also, some hypotheses must be made in order to complete the
simulation.
%
\citet{forgan_spatiotemporal_2011} argues that the times at which
different civilizations become intelligent follow a Gaussian
distribution, and then the distribution of inter-arrival times is an
inverse exponential.
%
We chose to assume that the distribution of the times of A events is a
stationary Poisson process, and then the distribution of the times
between the appearance of new consecutive \cetis{} is exponential.
%
Regarding the duration of a \ceti{}, we propose that the distribution
of the duration of \cetis{} is a stationary exponential distribution.
%
\citet{balbi_impact_2018} stress the fact that it would be desirable
to arrive at a theoretical statistical distribution of the lifespan of
civilizations, preferentially on the basis of the underlying
astrophysical and biological processes.
%
\citet{maccone_lognormals_2014} argues that this distribution should
be a log normal.  (EXPAND)

There is no a temporal window for the model, i.e., the simulation
starts when the process is already stable, and ends before any
galactic effect could modify or affect of fixed variables.
%
The probability for the rise of a \ceti{} is homogeneous over the GHZ.
%
Although the Galaxy has a well known spiral structure, the assumed
sparsity of civilizations makes this assumption very reasonably.
%
Otherwise, if the distribution of civilizations is not sparse, it
could be the case that the spiral arms would host most of the
civilizations, and contacts are frequent between closely located
civilizations.
%
In such a case we could expect that the interpretation of our results
could be considered as pessimist for the establishment of a contact
with our planet.
%
We also limit the possibilities of life or other types of civilizations
to the hypothesis usually stated for the definition of the GHZ
\citep{dayal_habitability_2016, gonzalez_galactic_2001,
lineweaver_galactic_2004, gonzalez_habitable_2005,
morrison_extending_2015, haqq-misra_evolution_2019,
rahvar_cosmic_2016, gobat_evolution_2016, rahvar_cosmic_2016}.
%
\citet{vukotic_astrobiological_2012} propose probabilistic cellular
automata model to explain the astrobiological history of the Milky
Way.
%
This means that we set aside possible civilizations that could survive
in severe conditions or unstable systems, which would prevent the
appearance of life as we know it.

While the aforementioned assumptions are basic conditions for most of
the stochastic processes observed in nature, we need to make stronger
assumptions related to the nature of the message.
%
The most simple assumption about the message itself, is that it
travels at light speed.
%
With this, we are considering messages sent through, for example,
electromagnetic radiation or even gravitational radiation, but we set
apart messages sent with mechanical means or physical objects, or
through some unknown technology that violates the known so far laws of
physics.
% 
For the communication of \cetis{} through messages sent isotropically,
we assume that the capacity to emit signals and to receive signals
occur at the same time.
%
This means that the ability to find another civilization develops at
the same time that the ability and intention to emit a signal in all
direction, intended to be detected by an unknown civilization.
%
Although there are several reasons to think that this could not be the
case, at large time scales it can be considered that both abilities
occur roughly at coincident epochs.
%
Another essential assumption is that all \cetis{} use the same signal
power, so that there is a maximum distance out to which it can be
detected.
%
It is worth mentioning that we are considering in our experimental
setup a system composed by several emitters and receivers across the
galaxy, under the hypothesis proposed before.
%
In such a system we compute probabilities of a \ceti{} making contact
with another \ceti{}, and the results must be interpreted as for a
randomly chosen \ceti{}, given certain conditions of homogeneity on an
ensemble of equivalent galaxies.
%
We stress the fact that we are not centered on the probabilities for
the Earth itself, but on a generalized \ceti{}.
%
Regarding the extent of the signals, we know that the distance from
which a signal from Earth could be detected using the current
technology is of a few hundreds light years, given that the signal was
sent to a specific direction.
%
%https://setiathome.berkeley.edu/forum_thread.php?id=80585&postid=1833000#1833000]
%https://www.engadget.com/2017/05/25/listening-to-starlight-our-ongoing-search-for-alien-intelligenc/]
%[from https://en.wikipedia.org/wiki/Fermi_paradox]
%
It is straightforward to propose and implement a distribution of
maximum distances, although this would increase the model complexity
at the cost of a larger uncertainty or of another dimension on the
hypothesis space.
%
%This proposal is based on a pronoid scenario, and we do not consider
%in this work the occurrence of paranoid or partially paranoid
%civilizations.
%%
%If that would be the case, the results obtained here can be taken as
%upper limits.

@@

The power law and exponential statistical distributions are among the
most common patterns found in natural phenomena.
%
For example, the distribution of the frequency of words in many
languages is known to follow a Zipf law (which is a power law).
%
Zipf law also describes population ranks of cities in various
countries, corporation sizes, income rankings, ranks of number of
people watching the same TV channel (Zipf's law).
%
% CITAR ALGO DE LA LEY DE ZIPF
%
The magnitudes of earthquakes, hurricanes, volcanic eruptions and
floods; the sizes of meteorites or the losses caused by business
interruptions from accidents, are also well described by power laws
(Sornette 2006).
%
The power law behavior has been observed in a variety of systems,
including for example stock market fluctuations, sizes of computer
files and word frequency in languages \citep{mitzenmacher_brief_2004,
newman_power_2005, simkin_theory_2006}
%
Power laws have also been widely used in biological sciences, e.g., in
analyzing connectivity patterns in metabolic networks
\citep{jeong_large_2000} and in the number of species observed per
unit area in ecology \citep{martin_origin_2006, frank_common_2009}.
% 
More examples can be found in the literature
\citep{martin_origin_2006, maccone_KLT_2010, barabasi_scale_2009,
maccone_evolution_2014, maccone_lognormals_2014,
benguigui_classificacion_2016}

In this work, the exponential distribution of lifespan and waiting
times is justified by considering the hypothesis that the process of
appearance of life in the galaxy is homogeneous and stationary.
%
That is, there is no preferred location within the GHZ for the
spontaneous appearance of life, and the emergence of a \ceti{} is
independent of the existence of previous \cetis{} in the galaxy.
%            
This is equivalent to proposing a Poisson process for the emergence of
\cetis{}, since there is a close relation between the number of events
in time or space and the waiting time or separation, respectively
\citep[e.g., ][]{ross_simulation_2012}.
%
That is, these are two alternative approaches to describing the same
process, a Poisson distribution for the number of events implies an
exponential distribution for their separations, and vice versa.
%
It should be emphasized that the exponential laws used in this work
are assumed as part of the working hypothesis, and instead of
analyzing results from a particular parameter chosen ad--hoc, we
explore the hypothesis space and analyze the impact of the values of
these parameters on the results.
%}}}

@@

The odds of a causal contact between two \cetis{} should not be
considered as the odds of a contact between two intelligent
civilizations, and in fact the later could be much lesser than the
former.
%
Indeed, in order to establish a contact between any two entities, a
minimum degree of compatibility must be accomplished without any
previous agreement, making the possibility of a contact with a message
that could be deciphered highly rare.
%
There is a trade--off between the simplicity and the complexity of the
experiments, and further analysis could be performed following this
framework in order to explore possible implications of the results for
more detailed configurations.
%
For example, the communication method (isotropic, collimated,
serendipitous) can affect the observables, making it necessary to
implement a correction factor.
%
Taking this into account, our results regarding the probability of
causal contact should be considered as upper limits for effective
contacts, since they depend on the efficiency of both the emitter and
the receiver to broadcast and scan the sky for intelligent signals,
respectively.
%
A correction by coverage ratio in the detection and by a targeting
ratio in the emission could be easily implemented in the simulation,
although the effect of reducing the probability of contact is
basically the product of the efficiency rations and thus such
implementation is not necessary.
%
Therefore, the values of the probabilities could be modified by a
constant correction factor equal to the combined emission/reception
efficiency \citep{smith_broadcasting_2009, anchordoqui_upper_2019,
forgan_collimated_2014}.
%
Other considerations include the effects of alignments, the use of
stars as sources or amplifiers, the nature of the message carrier, the
use of a spatial distribution that actually resembles the spiral shape
and the width of the disc of the Galaxy, different distribution
functions for the mean lifetime of \cetis{}, and different
efficiencies for the sphere of the causal region of each \ceti{}
resembling different searching strategies
\citep{hippke_interstellar_2017}.
%
It is also possible to consider that D$_{max}$ is different for
different \cetis{}.  For example a power law where powerful emissions
are rare and weak emissions are common could be an improvement to the
model.
%
In this work we chose not to implement this for the sake of
simplicity.
%
Finally, the role of the message contents could influence on the
lifespan of a \ceti{} that receives a message, although the
implementation of such behaviur would increase the number of free
parameters and would be highly speculative.

>>>>

Simulations are suitable tools to analyze systems that evolve with
time and involve randomness.
%
An advantage of simulations compared to theoretical approaches, is
that the former usually require less assumptions and simplifications,
and can be applied to systems where a theoretical model can not be
found.
%
In particular, many complex stochastic processes that can be described
by the evolution of the state of a system, can be efficiently modeled
with the discrete--event (hereafter, DE) simulation approach.
%
A system described with the DE paradigm is characterized by a set of
actors and events, where actors interact causally through a series of
events on a timeline and process these events in chronological order
\citep{ptolemaeus_system_2014, chung_simulation_2003,
ross_simulation_2012}.
%
Each event yields modifications to the values of the variables that define
the process, producing the corresponding change in the state of the
simulated system.
%
This method is well suited for the particular case of the diffusion of
intelligent signals in the galaxy, and allows to explore several
models easily.
%
We simulate the statistical properties of a set of points in space and
time that have a causal connection at light speed, hereafter dubbed
''Causal Contact Nodes'' (hereafter, \ceti{}).
%
We chose this generic name in order to stress the fact that in this
analysis no broadcasting or lookout activities are required, and
therefore not actual messages are considered, only the causal contact.
%
For the special case of a fully efficient node that emits and receives
isotropically, a causal contact node can be considered as a
broadcasting station that has the ability to detect signals through an
active lookout program.
%
Also, it is worth mentioning that this is a general approach, and not
necessarily a \ceti{} is the host of an intelligent civilization.
%
For instance, it can be associated with a planet where life has developed, became
intelligent, reached the skills required to find the right
communication channel, sustain a search and establish a contact.
%
Alternative message processing entities could be considered, for
example interstellar beacons where life has ceased to exist but
continue with its emission, or communication stations established by
probes or left intentionally by intelligent beings \citep[see, e.g.,
][]{peters_outer_2018, barlow_galactic_2013}.
%
In principle, these strategies could affect our results since it would
be easier to configure a cluster of \cetis{} that spread in time.
%
However, we do not consider these highly speculative alternatives at
this point.
%
For the purpose of this analysis, only the communication capability is
relevant, since we study the causal contacts between the locations.
%
The system is defined by a number of actors that represent \cetis{}
and appear at different instants in time, generating events that
produce meaningful changes in the variables that describe the system,
i.e., in the arrangement of \cetis{} and their network of causal
contacts.
%
For example, the appearance of a new \ceti{} in a region which is
filled with a signal emitted by another node, will increase the
number of active \cetis{} and the number of pairs of \cetis{} in
causal contact.
%
Assuming some simple hypotheses, the discrete events method can be
performed taking into account a small number of variables, which allow
to analyze the variation of the results in the model parameter space.

@@

\ffn{1} We illustrate in the Fig.~\ref{F_scheme} the schematic
representation of the region causally connected to a given \ceti{}.
%
To this end, we use space--time diagrams, where time is represented on
the horizontal axis, and space is represented in the vertical axis.
%
For the sake of simplicity, we show in the plot the light travel
distance, i.e., the distance traversed by any signal spreading from
the emitter at light speed (for example, any electromagnetic signal).
%
In this scheme, a light pulse would follow a trajectory represented by
a line at 45 degrees from the axes, given the units of the time and
space axes are years and light years, respectively.
%
Panel (a) of the Figure represents the region $R$ (shaded polygon) in
space and time which is causally connected to the emitter (left
vertex), following an event (hereafter A type event) in which the node
acquires the communication capability, or becomes ''active''.
%
The sphere of first contact (SFC) is a sphere centered on the emitter,
represented in the Figure by the left angle of the region $R$.
%
This sphere grows until its radius reaches the D$_{max}$ distance, in
which it would be no longer detectable due to the decrease in the
energy per unit area, which falls under the assumed detectability
threshold.
%
Similarly, the surface of last contact (SLC) is another sphere that
grows from an event in which the nodes ceases to possess the
communication capability (D type event), and carries the last signal
produced by the emitter.
%
The region which is causally connected to the emitter is then limited
by these two spheres, and has the shape of a sphere (before the D
event) or of a spherical shell (after the D event).
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b) of the Figure.
%
In this scheme, the entire lifetime of a \ceti{} can then be
represented by a polygon, limited in the spatial axis by the maximum
distance of the signal $D_{max}$ and in the time axis by the time span
between the starting time (A event, concave vertex) and the ending
time (D event, convex vertex).
%
This representation allows to easily visualize the important events
that result from the existence and communications of two \cetis{}.
%
The points in time where the \cetis{} acquire their communicating
capacity are dubbed ''Awakening'' events ($A_1$ and $A_2$ in the
Figure).
%
Similarly, the points in time where the \cetis{} loss their
communicating capacity are dubbed ''Doomsday'' events ($D_1$ and $D_2$
in the Figure).
%
The points in space--time where the first contact is produced for each
one of the \cetis{}, are defined as ''Contact'' events, shown as $C_1$
and $C_2$ in the Figure.
%
Finally, the two points where the contact is lost for each one the
\cetis{}, are denominated ''Blackout'' events, shown as $B_1$ and
$B_2$ in the Figure.
%
The receiver node E$_1$ can listen signals from the emitter node
E$_2$, from the first contact event (t=C$_{12}$) up to the last
contact event (t=B$_{12}$).
%
Similarly, the receiver node E$_2$ can listen signals from the emitter
node E$_1$, from t=C$_{21}$ up to t=B$_{21}$.
%
The time intervals for the open communication channel are determined
by the ''C'' and ''B'' type events.
%
It is important to notice that there is a time delay between the
contact events of the two \cetis{} involved in this analysis, and also
a time delay between the two blackout events, so that the time range
when a bidirectional contact is possible occurs between the maximum
time of the contact points and the minimum time between the blackout
points.

@@

The temporal structure that emerges naturally from the experimental
setup implies that,
as a fundamental property of the simulations,
a causal connection can be produced without the need of the two
involved locations being concurrent or active at the same time.
%
This property of the system arises as a
consequence of the large spatial and temporal scales, where a message
is transmitted at the (relatively small) light speed.
%
Although a \ceti{} could be active for enough time so as to transmit a
message at large distances, the limited power of the message and the
dilution that runs as the squared distance from the source, imposes a
detectability limit.
%
As a consequence of this limitation and of their finite lifetime,
considered as the period of time between the acquisition and loss of
communicating capacity, each \ceti{} will fill a spherical shell
region of the galaxy, limited by two concentric spherical surfaces.
%
The leading front, or surface of first contact (SFC) grows from the
central \ceti{} until it reaches the maximum distance $D_{max}$.
%
Thus, the volume reached causally by a \ceti{} is initially a growing
sphere.
%
Following the end of the civilization, there is still a region which
is filled with the emitted signals.
%
This approach has been also considered other statistical models
\citep[e.g., ][]{smith_broadcasting_2009, grimaldi_signal_2017, Grimaldi2018}.
%
The trailing front, or surface of last contact (SLC) also grows from
the central \ceti{}, with a delay with respect to the SFC equivalent
to the lifetime of the node, and produces a spherical shell region.
%
Any other \ceti{} within this region will be aware of the originating
node, even if it has disappeared before the time of contact.
%
This region will grow if the surface of first contact has not yet
reached the maximum distance $D_{max}$, and will shrink otherwise
until the surface of last contact reaches $D_{max}$, producing as a
result the loss of all signal from the central \ceti{}.
%
In our approach, we consider a model galaxy where the width of the
disk is negligible with respect to the radius of the disk.
%
In the 2D simulation only the intersection of the communicating
spherical shells with the plane of the galaxy is relevant, and produce
the corresponding circles or rings for the filled spheres or annular
regions of the spherical shells, respectively.
%
The initially growing communicating sphere, is shown over space--time
diagrams, where time is represented on the horizontal direction, and
space is represented in the vertical direction.
%
In the Fig.~\ref{F_scheme} the two
emitters in panel (b), E$_1$ and E$_2$, reach each other at different times.
%
The time span for E$_i$ is (A$_i$ , D$_i$) , for $i = {1, 2}$.
%
Emitter $i$ can listen to emitter $j$ between C$_{ij}$ and B$_{ij}$.
%
The type and length of causal contact in both directions depend on the
distance and time lag between the awakening events, the maximum
distance that a signal can reach and the time period in which each
emitter is active.
%
\ffn{2}
%
In the Fig.~\ref{F_messages} we show a schematic representation of the
possible cases in which a \ceti{} E can be in causal contact with a
\ceti{} R.
%
The duration of the causal contact in one direction depends on several
factors, mainly the lifetimes, the time period between the A events
and the distance between the two nodes involved in the causal
contact.
%
As it can be seen, the direction of the causal contact does not imply
any ordering for the A events of the involved nodes.
%
In the implementation of the DE simulation, the six cases shown in the
Figure determine the succession of events.
%
Each time an event is produced, the list of upcoming events is updated
according to these cases.

In out experimental configuration, there is no a temporal window for
the model, i.e., the simulation starts when the process is already
stable, and ends before any galactic effect could modify or affect of
fixed variables.
%
Also, we assume that the probability for the rise of a \ceti{} is
homogeneous over the GHZ.
%
Although the Galaxy has a well known spiral structure, the assumed
sparsity of nodes (intelligent civilizations) does make this
assumption fairly reasonable.
%
Otherwise, if the distribution of nodes is not sparse, it could be the
case that the spiral arms would host most of the \cetis{}, and
contacts are frequent between closely located nodes.
%
In such a case we could expect that the interpretation of our results
could be considered as pessimist for the establishment of a contact
with our planet.
%
We also limit the possibilities of life or other types of
civilizations to the hypothesis usually stated for the definition of
the GHZ \citep{dayal_habitability_2016, gonzalez_galactic_2001,
lineweaver_galactic_2004, gonzalez_habitable_2005,
morrison_extending_2015, haqq-misra_evolution_2019,
rahvar_cosmic_2016, gobat_evolution_2016, rahvar_cosmic_2016}.
%
For example, \citet{vukotic_astrobiological_2012} propose
probabilistic cellular automata model to explain the astrobiological
history of the Milky Way.
%
This means that we set aside possible civilizations that could survive
in severe conditions or unstable systems, which would prevent the
appearance of life as we know it.

While the aforementioned assumptions are basic conditions for most of
the stochastic processes observed in nature, we need to make stronger
assumptions related to the nature of the message.
%
The most simple assumption about the message itself, is that it
travels at light speed.
%
With this, we are considering messages sent through, for example,
electromagnetic radiation or even gravitational radiation or neutrino
communication \citep{wright_how_2018}, but we set
apart messages sent with mechanical means or physical objects
\citep[e.g., ][]{Armstrong2013}, or through some unknown technology
that violates the known laws of physics.
% 
For the communication of \cetis{} through messages sent isotropically,
we assume that the capacity to emit signals and to receive signals
occur at the same time.
%
This means that the ability to find another civilization develops at
the same time that the ability and intention to emit a signal in all
directions, intended to be detected by an unknown civilization.
%
Although there are several reasons to think that this could not be the
case, at large time scales it can be considered that both abilities
occur roughly at coincident epochs.
%
Another essential assumption is that all \cetis{} use the same signal
power, so that there is a maximum distance out to which it can be
detected.
%
It is worth mentioning that we are considering in our experimental
setup a system composed by several emitters and receivers across the
galaxy, under the hypothesis proposed before.
%
In such a system we compute probabilities of a \ceti{} making contact
with another node, and the results must be interpreted as for a
randomly chosen \ceti{}, given certain conditions of homogeneity on an
ensemble of equivalent galaxies.
%
Then, we stress the fact that we are not centered on the probabilities
for the Earth itself, but on a generalized \ceti{}.
%
Regarding the extent of the signals, we know that the distance from
which a signal from Earth could be detected using the current
technology is of a few hundreds light years, given that the signal was
sent to a specific direction.
%
%https://setiathome.berkeley.edu/forum_thread.php?id=80585&postid=1833000#1833000]
%https://www.engadget.com/2017/05/25/listening-to-starlight-our-ongoing-search-for-alien-intelligenc/]
%[from https://en.wikipedia.org/wiki/Fermi_paradox]
%
It is straightforward to propose and implement a distribution of
maximum distances, although this would increase the model complexity
at the cost of a larger uncertainty or of another dimension on the
hypothesis space.
%
%This proposal is based on a pronoid scenario, and we do not consider
%in this work the occurrence of paranoid or partially paranoid
%civilizations.
%%
%If that would be the case, the results obtained here can be taken as
%upper limits.

@@

The temporal structure of the process is defined by two distribution
parameters that represent the mean time interval that a \ceti{}
can emit and receive signals, and the mean time interval
between the emergence of consecutive \cetis{}.
%
The spatial structure of the simulation is given by the size and shape
of the Galactic Habitable Zone (GHZ) and the maximum distance a signal can
travel to be detected (D$_{max}$).
%
The parameters for the temporal distributions also determine the
spatial properties, since the density of active \cetis{} in the galaxy
depend on these two parameters.
%
For example, a small $\tau_a$ and a large $\tau_s$ will produce a
galaxy densely populated with \cetis{}.
%
Also, some hypotheses regarding the shapes of the distributions of the
temporal parameters must be made in order to complete the
simulation.
%
\citet{forgan_spatiotemporal_2011} argues that the times at which
different civilizations become intelligent follow a Gaussian
distribution, and then the distribution of inter-arrival times is an
inverse exponential.
%
We choose to assume that the distribution of the times of A events is a
stationary Poisson process, and then the distribution of the times
between the appearance of new consecutive \cetis{} is exponential.
%
Regarding the duration of a \ceti{}, we propose that the distribution
of the duration of \cetis{} is a stationary exponential distribution.
%
\citet{balbi_impact_2018} stress the fact that it would be desirable
to arrive at a theoretical statistical distribution of the lifespan of
civilizations, preferentially on the basis of the underlying
astrophysical and biological processes.
%
\citet{maccone_lognormals_2014} argues that this distribution should
be a log normal.

The power law and exponential statistical distributions are among the
most common patterns found in natural phenomena.
%
For example, the distribution of the frequency of words in many
languages is known to follow a Zipf's law (which is a power law).
%
This distributions arise any time a phenomena is characterized by
commonly occurring small events and rarely produced large events
\citep[e.g. ][]{adamic_zipf_2000}.
%
Zipf law also describes population ranks of cities in various
countries, corporation sizes, income rankings, ranks of number of
people watching the same TV channel.
%
The magnitudes of earthquakes, hurricanes, volcanic eruptions and
floods; the sizes of meteorites or the losses caused by business
interruptions from accidents, are also well described by power laws
\citep{sornette_critical_2006}.
%
The power law behavior has been observed in a variety of systems,
including for example stock market fluctuations, sizes of computer
files and word frequency in languages \citep{mitzenmacher_brief_2004,
newman_power_2005, simkin_theory_2006}
%
Power laws have also been widely used in biological sciences, e.g., in
analyzing connectivity patterns in metabolic networks
\citep{jeong_large_2000} and in the number of species observed per
unit area in ecology \citep{martin_origin_2006, frank_common_2009}.
% 
More examples can be found in the literature
\citep{martin_origin_2006, maccone_KLT_2010, barabasi_scale_2009,
maccone_evolution_2014, maccone_lognormals_2014,
benguigui_classificacion_2016}

In this work, the exponential distribution of lifespan and waiting
times is justified by considering the hypothesis that the process of
appearance of life in the galaxy is homogeneous and stationary.
%
That is, there is no preferred location within the GHZ for the
spontaneous appearance of life, and the emergence of a \ceti{} is
independent of the existence of previous nodes in the galaxy.
%            
This is equivalent to proposing a Poisson process for the emergence of
\cetis{}, since there is a close relation between the number of events
in time or space and the waiting time or separation, respectively
\citep[e.g., ][]{ross_simulation_2012}.
%
That is, these are two alternative approaches to describing the same
process, a Poisson distribution for the number of events implies an
exponential distribution for their separations, and vice versa.
%
It should be emphasized that the exponential laws used in this work
are assumed as part of the working hypothesis, and instead of
analyzing results from a particular parameter chosen ad--hoc, we
explore the hypothesis space and analyze the impact of the values of
these parameters on the results.
%}}}

@@

In this Section we discuss the degree of complexity in the model,
considering a compromise between the accuracy of the model and the
number of parameters that are free or with a high uncertainty.
%
Firstly, we emphasize that the odds of a causal contact between two
\cetis{} should not be considered as the odds of a contact between two
intelligent civilizations, and in fact the later could be much lesser
than the former.
%
Indeed, in order to establish a contact between any two entities, a
minimum degree of compatibility must be accomplished without any
previous agreement, making the possibility of a contact with a message
that could be deciphered highly rare.
%
Besides the trade--off between the simplicity and the complexity of
the experiments, further analysis could be performed following this
framework in order to explore possible implications of the results for
more detailed configurations.
%
For example, the communication method (isotropic, collimated,
serendipitous) can affect the observables, making it necessary to
implement a correction factor.
%
Taking this into account, our results regarding the probability of
causal contact should be considered as upper limits for effective
contacts, since they depend on the efficiency of both the emitter and
the receiver to broadcast and scan the sky for intelligent signals,
respectively \citep{grimaldi_signal_2017}.
%
A correction by coverage ratio in the detection and by a targeting
ratio in the emission could be easily implemented in the simulation,
although the effect of reducing the probability of contact is
basically the product of the efficiency ratios and thus such
implementation is not necessary.
%
Therefore, the values of the probabilities could be modified by a
constant correction factor equal to the combined emission/reception
efficiency \citep{smith_broadcasting_2009, anchordoqui_upper_2019,
forgan_collimated_2014} or for beam--like transmissions
\citep{grimaldi_signal_2017}.
%
Other considerations include the effects of alignments, the use of
stars as sources or amplifiers \citep{Edmondson2003, borra_searching_2012}, 
the nature of the message carrier, the
use of a spatial distribution that actually resembles the spiral shape
and the width of the disc of the Galaxy, different distribution
functions for the mean lifetime of \cetis{}, and different
efficiencies for the sphere of the causal region of each \ceti{}
resembling different searching strategies
\citep{hippke_interstellar_2017}.
%
It is also possible to consider that D$_{max}$ is different for
different \cetis{}.  For example a power law where powerful emissions
are rare and weak emissions are common could be an improvement to the
model.
%
In this work we chose not to implement this for the sake of
simplicity.
%
Finally, the role of the message contents could influence on the
lifespan of a \ceti{} that receives a message, although the
implementation of such behavior would increase the number of free
parameters and would be highly speculative.
